---
title: "Homework 2: Classification"
subtitle: |
  | MACS 30100: Perspectives on Computational Modeling
  | University of Chicago
author: Wanxi Zhou
output: pdf_document
---

## A Theoretical Problem

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(MASS)

SEED = 1234
K = 1000
N = 1000

# Helper function for classification
bayes_decision <- function(x1, x2){
  y = x1 + x1 ^2 + x2 + x2 ^ 2
  return(ifelse(y >= 0, TRUE, FALSE))
}

# Helper function for data set generation
create_dataset <- function(n, seed){
  set.seed(seed)
  return(tibble(x1 = runif(n, min = -1, max = 1),
                x2 = runif(n, min = -1, max = 1),
                y = bayes_decision(x1, x2)))
}

# Helper function for error rate calculation
err <- function(mod, err_type){
  return(sum(err_type$y != predict(mod, err_type)$class) / nrow(err_type))
}

lda_train_err = as.numeric(N)
qda_train_err = as.numeric(N)
lda_test_err = as.numeric(N)
qda_test_err = as.numeric(N)

# Compute training and test errors of the 2 methods for 1000 simulations
for(i in 1:K){
  dataset <- create_dataset(N, SEED)
  split <- initial_split(dataset,
                         prop = 0.8)
  train <- training(split)
  test <- testing(split)
  lda_mod <- lda(y ~ ., data = train)
  qda_mod <- qda(y ~ ., data = train)
  lda_train_err[i] <- err(lda_mod, train)
  qda_train_err[i] <- err(qda_mod, train)
  lda_test_err[i] <- err(lda_mod, test)
  qda_test_err[i] <- err(qda_mod, test)
  SEED <- SEED + 1
}

err_tb <- tibble(error_rate = c(lda_train_err, qda_train_err,
                       lda_test_err, qda_test_err),
                 error_type = c(rep("lda_train_err", N),
                       rep("qda_train_err", N),
                       rep("lda_test_err", N),
                       rep("qda_test_err", N)))

# Draw box plot for comparing the training and test errors
# of the 2 methods
myColors <- ifelse(levels(err_tb$error_type) == "lda_train_err" |
                   levels(err_tb$error_type) == "lda_test_err", 
                   rgb(0.1,0.1,0.7,0.5), rgb(0.8,0.1,0.3,0.6))
err_tb %>%
  ggplot(aes(x = error_type, y = error_rate, fill = error_type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("lightblue", "lightblue",
                               "orange", "orange"))

# Generate summary table
err_tb %>%
  group_by(error_type) %>%
  summarize(mean = mean(error_rate),
            median = median(error_rate),
            var = var(error_rate),
            sd = sd(error_rate),
            max = max(error_rate),
            min = min(error_rate),
            .groups = "drop")
```


*From both the box plot and the table, we could conclude that QDA outperforms
LDA when the Bayes decision boundary is non-linear (especially of the quadratic
form). Both the training and testing errors of the QDA models are less than the
LDA classifier and the former also has lower variance, which means that QDA
could produce more accurate model for prediction in this case.*

*For both approaches,their training error and testing error share similar value, 
with the testing error slightly lower than the training error. As expected, it 
could be that the model is more familiar with the training set since it is 
trained on that set.*

*In non-linear contexts like this, since QDA generates a quadratic decision
boundary, it could more accurately approximate the real Bayes classifier,
resulting in lower bias in estimates. Therefore, even though higher flexibility 
raises the variance of the QDA model, lower biases guarantee better performance
of QDA than LDA under this circumstance.*


## An Applied Problem

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(tidymodels)
library(MASS)
library(class)
library(klaR)
library(plotROC)
library(gridExtra)

 # Load and preprocess the data set
anes <- read_csv(here("data", "anes_pilot_2016.csv"))
anes_sample <- anes %>%
  dplyr::select(fttrump, ftobama, fthrc, ftrubio, ideo5, pid3) %>%
  mutate(fttrump = replace(fttrump, fttrump > 100, NA),
         ftobama = replace(ftobama, ftobama > 100, NA),
         fthrc = replace(fthrc, fthrc > 100, NA),
         ftrubio = replace(ftrubio, ftrubio > 100, NA),
         pid3 = replace(pid3, pid3 > 3, NA),
         ideo5 = replace(ideo5, ideo5 > 5, NA),
         democrat = as.factor(ifelse(pid3 == 1, 1, 0))) %>%
  drop_na() %>%
  dplyr::select(-pid3) %>%
  relocate(democrat)

# Split the data set and use the training set for cross validation approach
set.seed(222)

split <- initial_split(anes_sample,
                       prop = 0.8)
train <- training(split)
test <- testing(split)
cv_train <- vfold_cv(data = train,
                     v = 10)

# Create recipe and models (except KNN)
recipe <- recipe(democrat ~ ., 
                 data = anes_sample) 

mod_logistic <- logistic_reg() %>% 
  set_engine("glm") %>% 
  set_mode("classification")

mod_lda <- discrim::discrim_linear() %>%
  set_engine("MASS")

mod_qda <- discrim::discrim_regularized(frac_common_cov = 0) %>%
  set_engine("klaR") %>%
  translate()


########################## LOGISTIC REGRESSION ###############################
# Start work flow
workflow <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(mod_logistic)

# Collect test error and AUC
res_logistic <- workflow %>%
  fit_resamples(resamples = cv_train) %>%
  collect_metrics() %>%
  as_tibble() %>%
  dplyr::select(.metric, mean)

res_logistic[1, ] = list("error",
                         as.numeric(1 - res_logistic[1, 2]))
res_logistic <- res_logistic %>%
  add_column(model_type = "logistic")

# Plot ROC curve
predictions = NULL
truths = NULL
id = NULL
for(i in 1:10){
  fold <- cv_train$splits[[i]]
  tr <- analysis(fold)
  te <- assessment(fold)
  train_mod <- glm(democrat ~ .,
                   data = tr,
                   family = "binomial")
  pred <- predict(train_mod,
                  newdata = te,
                  type = "response")
  truths = c(truths,
             as.numeric(as.character(te$democrat)))
  predictions = c(predictions, pred)
  id = c(id, rep(as.character(fold$id), nrow(te)))
}

roc_logistic <- tibble(id = id,
                       truth = truths,
                       pred = predictions) %>%
  ggplot(aes(m = pred, d = truth, color = id)) +
  geom_roc(labels = F, size = 0.5) +
  ggtitle("Logistic ROC Curve") +
  scale_color_manual(values = c(rep("grey", 10))) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.3))

#################################### LDA #####################################
# Change model in work flow
workflow <- update_model(workflow, mod_lda)

# Collect test error and AUC
res_lda <- workflow %>%
  fit_resamples(resamples = cv_train) %>%
  collect_metrics() %>%
  as_tibble() %>%
  dplyr::select(.metric, mean)

res_lda[1, ] = list("error",
                     as.numeric(1 - res_lda[1, 2]))
res_lda <- res_lda %>%
  add_column(model_type = "lda")

# Plot ROC curve
predictions = NULL
truths = NULL
id = NULL
for(i in 1:10){
  fold <- cv_train$splits[[i]]
  tr <- analysis(fold)
  te <- assessment(fold)
  train_mod <- lda(democrat ~ .,
                   data = tr)
  pred <- predict(train_mod,
                  newdata = te)$posterior[,2]
  truths = c(truths,
             as.numeric(as.character(te$democrat)))
  predictions = c(predictions, pred)
  id = c(id, rep(as.character(fold$id), nrow(te)))
}

roc_lda <- tibble(id = id,
                  truth = truths,
                  pred = predictions) %>%
  ggplot(aes(m = pred, d = truth, color = id)) +
  geom_roc(labels = F, size = 0.3) +
  ggtitle("LDA ROC Curve") +
  scale_color_manual(values = c(rep("grey", 10))) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.3))

#################################### QDA #####################################
# Change model in work flow
workflow <- update_model(workflow, mod_qda)

# Collect test error and AUC
res_qda <- workflow %>%
  fit_resamples(resamples = cv_train) %>%
  collect_metrics() %>%
  as_tibble() %>%
  dplyr::select(.metric, mean)

res_qda[1, ] = list("error",
                     as.numeric(1 - res_qda[1, 2]))
res_qda <- res_qda %>%
  add_column(model_type = "qda")

#Plot ROC curve
predictions = NULL
truths = NULL
id = NULL
for(i in 1:10){
  fold <- cv_train$splits[[i]]
  tr <- analysis(fold)
  te <- assessment(fold)
  train_mod <- qda(democrat ~ .,
                   data = tr)
  pred <- predict(train_mod,
                  newdata = te)$posterior[,2]
  truths = c(truths,
             as.numeric(as.character(te$democrat)))
  predictions = c(predictions, pred)
  id = c(id, rep(as.character(fold$id), nrow(te)))
}

roc_qda <- tibble(id = id,
                  truth = truths,
                  pred = predictions) %>%
  ggplot(aes(m = pred, d = truth, color = id)) +
  geom_roc(labels = F, size = 0.3) +
  ggtitle("QDA ROC Curve") +
  scale_color_manual(values = c(rep("grey", 10))) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))

#################################### KNN #####################################
# Select best k value for KNN
tb_knn <- tibble(k = 1:10,
                 err_rate = NA,
                 auc = NA)

for(i in 1:10){
  mod_knn <- nearest_neighbor(neighbors = i,
                              dist_power = 2) %>%
    set_engine("kknn") %>%
    set_mode("classification")
  
  workflow <- update_model(workflow, mod_knn)

  res_knn <- workflow %>%
    fit_resamples(resamples = cv_train) %>%
    collect_metrics()

  tb_knn$err_rate[i] <- res_knn$mean[1]
  tb_knn$auc[i] <- res_knn$mean[2]
}

# Select k = 10
print(tb_knn)

# Collect test error and AUC
res_knn <- workflow %>%
  fit_resamples(resamples = cv_train) %>%
  collect_metrics() %>%
  as_tibble() %>%
  dplyr::select(.metric, mean)

res_knn[1, ] = list("error",
                     as.numeric(1 - res_knn[1, 2]))
res_knn <- res_knn %>%
  add_column(model_type = "knn")

# Plot ROC curve
predictions = NULL
truths = NULL
id = NULL
for(i in 1:10){
  fold <- cv_train$splits[[i]]
  tr <- analysis(fold)
  te <- assessment(fold)
  knn_mod <- knn(tr, te, tr$democrat,
                 k = 10, prob = T)
  pred <- knn(tr, te, tr$democrat, k = 10, prob = T)
  prop <- attributes(pred)$prob
  prop <- ifelse(pred == 0, 1 - prop, prop)
  truths = c(truths,
             as.numeric(as.character(te$democrat)))
  predictions = c(predictions, prop)
  id = c(id, rep(as.character(fold$id), nrow(te)))
}

roc_knn <- tibble(id = id,
                  truth = truths,
                  pred = predictions) %>%
  ggplot(aes(m = pred, d = truth, color = id)) +
  geom_roc(labels = F, size = 0.3) +
  ggtitle("KNN ROC Curve") +
  scale_color_manual(values = c(rep("grey", 10))) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))


## Aggregate test errors, ROC curves and AUC for model selection
grid.arrange(roc_logistic, roc_lda, roc_qda, roc_knn,
             ncol = 2, nrow = 2)

bind_rows(res_logistic, res_lda, res_qda, res_knn) %>%
  group_by(.metric) %>%
  arrange(desc(mean))


#################### SELECT BETWEEN LOGISTIC AND LDA ########################
cov(anes_sample[anes_sample$democrat == 0,2:6])
cov(anes_sample[anes_sample$democrat == 1,2:6])


############################ CHOOSE LOGISTIC ################################
# Evaluate logistic model using test set
final_mod <- workflow %>%
  update_model(mod_logistic) %>%
  last_fit(split)

# Collect test error and AUC
m <- final_mod %>%
  collect_metrics() %>%
  as_tibble() %>%
  dplyr::select(.metric, .estimate)

m %>%
  add_row(.metric = "err_rate",
          .estimate = 1 - m$.estimate[1])

# Generate confusion metric
res <- final_mod %>%
  collect_predictions()

res %>%
  conf_mat(truth = democrat, 
           estimate = .pred_class,
           dnn = c("Pred", "Truth"))

# Plot logistic regression predictions
res %>%
  ggplot() +
  geom_bar(aes(x = .pred_class, 
               fill = democrat)) +
  facet_wrap(~ democrat) +
  labs(title = "Logistic Regression Predictions",
       x = "Predicted Party Affiliations", 
       fill = "Democrat",
       caption = "Note: Facits are the truth") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = c("lightblue", "orange"))

# Plot ROC curve
tibble(truth = as.numeric(as.character(res$democrat)),
       pred = res$.pred_1) %>%
  ggplot(aes(m = pred, d = truth)) +
  geom_roc(labels = F) +
  ggtitle("Logistic ROC Curve") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

*According to the test errors, AUC, and ROC curves of different models
computed from the training data using cross validation, we can observe that
logistic regression and LDA models share almost identical performance
results. However, since each class of the sample data demonstrates variant
covariance matrices, logistic regression is favored in this context, for the
purpose of more accurately predicting future unfamiliar data sets.*

*According to the final evaluation results, the logistic model achieves an
approximately 0.2 test error and 0.9 AUC, exhibiting reliable prediction
performance. The ROC curve also hugs the top left corner, indicating a both low 
Type I error and Type II error. Therefor, the area under the curve approaches 1.*

*The relatively higher performance of logistic regression, along with LDA,
indicates the possibility that the decision boundary of party affiliation in 
this case might be approximately linear. Compared to all the other classifiers,
the selected approach, logistic regression, is less flexible, and consequently
has lower variance. Furthermore, if in reality, the response class is indeed
separated by a linear boundary, the logistic model could generate less biased
estimates and perform relatively accurate classification.*

*The result supports the theorical assumption that concepts (e.g., feelings
towards various political figures) indirectly related to one’s party affiliation
could actually drive their party affiliation.*
